%--------------------------------------------------------------------------------------------------
%
\chapter*{Acknowledgments}
\pdfbookmark[0]{Acknowledgments}{Acknowledgments}
%--------------------------------------------------------------------------------------------------

I would like to express great appreciation to my PhD supervisor Prof. Dunja
Mladenić and co-supervisor Prof. John Shawe-Taylor for their guidance and advice
throughout my studies. Special thanks go to Primož Škraba, for his invlauable advice and contributions to my work. 
I would also like to thank Marko Grobelnik for providing inspiration and encouragement for my research work.

Assistance provided by my collaborators Andrej Muhič, Blaž Fortuna, Gregor Leban 
and Sabrina Guettes was greatly appriciated. It was my great pleasure working with you.
I would also like to thank my friends from UCL: Tom Diethe, David Roi Hardoon and Zakria Hussain
for great company and discussions during my visits to UCL.

My gratitude goes to the members of my doctoral committee, Prof. Nada Lavrač, Prof.
Bor Plestenjak and Prof. Nicolò Cesa-Bianchi, for their valuable comments and
remarks.

Finally, I wish to thank Zala, my parents Darja and Erik and my sister Nika 
for their support and encouragement.

\rule{0.5\textwidth}{.4pt}

The author gratefully acknowledges that the funding for this work was provided by the projects
SMART (IST-5-033917-STP), X-LIKE (ICT-257790-STREP), MultilingualWeb (PSP-2009.5.2 Agr.\# 250500), 
TransLectures (FP7-ICT-2011-7), PlanetData (ICT-257641-NoE), RENDER (ICT-257790-STREP), XLime (FP7-ICT-611346), 
and META-NET (ICT-249119-NoE). 